Introducing an Ecological Trait-data Standard

### 2017-09-15 manuscript/manuscript.Rmd
update manuscript, insert figures and external links

---
title: "Introducing an Ecological Trait-data Standard"
author: "Florian D. Schneider, Nadja K. Simons, Andreas Ostrowski, Caterina Penone, Malte Jochum, Gaetane Le Provost"
geometry: margin=1.5in
bibliography: div05.bib
csl: amnat.csl
output:
  word_document: default
  pdf_document: default
header-includes: \usepackage{lineno}
fontsize: 12pt
---

@@

<!-- 
The aim of the methods paper is to 
- highlight the importance of data standards for functional traits
- provide standards and tools for decentralised implementation

1. importance of trait based ecology, current research questions, current applications of traits. 
2. Overview of existing database initiatives. Offering access management and harmonization;

(table 1 listing of open an restricted trait databases and datasets)

3. Heterogeneity of data standards and centralised databases. Open Access, publication in low-threshold data repositories becomes more important. Decentralisation of databases and importance of data brokering requires data standards. 
- lack of a universal traitdata standard.

1. minimal definition of a trait dataset
2. possible data formats: matrix, occurrence table, measurement long table 
- species--trait matrix is common for sharing taxon-level averages or facts, common for publishing species-level data and literature data. but not raw-data. The means of obtaining the average are intransparent and may be biased by local conditions. 
- occurrence table is common for empirical replicated measurements of traits, raw data. Ideal for correlation analysis of traits. Also, multivariate traits, e.g. morphometric landmarks or biochemical trais. Not ideal for merging data from multiple sources. 
- measurement table, longtable: ideal for raw data. Machine readable, simple to standardize and for sharing data, merging and combining data. 
- additional information may apply to different layers: on measurement or occurrence, co-variate analysis, environmental filters, spatial distribution.

(figure 1 species-trait matrix plus additional information, vs longtable plus additional information )

# Towards an Ecological Traitdata Standard

1. definition of structure: longtable format is most feasible to carry information at measurement level, link to external datasets via globally unique identifiers (or dataset specific)
- add occurrence level information (e.g. location, sex) and measurement level (e.g. aggregated data, statistics) information
- apply taxon name standardization
- apply trait standardization, link to definitions (trait lists, thesaurus) via globally unique identifiers, semantic network; apply unit and factor level harmonization

## A glossary of terms

- We provide: Glossary of terms for trait data and trait definitions

(table 2: most relevant terms of the traitdata standard, scheme of a term definitition)
(table 3: terms for trait definitions / trait lists)

## Computational tools for producing compliant data

- R package traitdataformat
- standardize and harmonize trait data for publication
- merge, split, apply trait data of different sources

(figure 2 process chart from original file to standardized output)

# Discussion

- With these tools, traitdata will be easier to harmonize 
- decentralised structures, not limited by infrastructural projects
- open source community development, invite contributions to traitdata standard and toolchain of R-package
- encourage development of trait definitions and semantic ontologies
- a global analysis of ecological traits will be facilitated

@@

Trait-datasets take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums. The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level (@laliberte10). Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (Refs).

@@

With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL)  TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC). We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility. The full list of terms of the traitdata standard including further explanation can be assessed at http://www.xxxxx.

@@

@@

In principle, transferring and harmonizing traitdata using the package is as simple as:

```r
library("traitdataform")
traitdataset1 <- standardize.traits(read.csv("path/to/data.csv"),
            thesaurus = as.thesaurus("http://url.of/thesaurus.csv"),
            taxa = "name_correct", units = "mm")
```

For a detailed description on how to use the package see the package vignette (). The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page ().

<!--More step-by-step examples here on how to create linked datasets, add a trait thesaurus, merge two datasets  -->

(figure 2 process chart from original file to standardized output)

>>>>

---
title: "Introducing an Ecological Trait-data Standard"
author: "Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Nadja K. Simons, Andreas Ostrowski, Caterina Penone"
csl: amnat.csl
output:
  word_document:
    keep_md: yes
  pdf_document: default
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

@@

Trait-datasets take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, they usually are reported in a species $\times$ trait matrix format, with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums. The matrix format is widely used for the production of lookup-tables at the species level, which for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level (@laliberte10). Also, missing information about the behaviour and functional role of species of little ecological record may be inferred from these tables (@Refs).

@@

With this framework, we build upon the structure proposed for the Encyclopedia of Life (EOL)  TraitBank [@parr16], which uses terms of the Darwin Core standard (DWC). We expand the definitions of these terms for the use case of trait data and add further terms to cover the special demands of trait-based research, while maintaining compatibility. The full list of terms of the traitdata standard including further explanation can be assessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication).

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.75\textwidth, angle=90]{figures/fig2.png}
\caption{Process chart of the functions provided within the R package.}

\end{figure}

@@

The package functions comprise a toolchain where each function can be piped as an input into the next. A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing traitdata as simple as:

```r
library("traitdataform")
traitdataset1 <- standardize(read.csv("path/to/data.csv"),
            thesaurus = as.thesaurus("http://url.of/thesaurus.csv"),
            taxa = "name_correct", units = "mm")
```

For a detailed description on how to use the package see the package vignette (https://github.com/fdschneider/traitdataform/raw/master/docs/vignette.pdf). The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).
